
\chapter*{Our Key Partners}\label{keypartners}

\vspace{-.5cm}

The organisers of the conference are indebted to the various institutions and individuals for the invaluable help rendered by them, without which this work would just not have been possible. It is a pleasure to thank them heartily for the same.

We are indebted to our individual and institutional sponsors: Sri Mohandas Pai and Foundation for Indian Civilization Studies, Sri J K Jhaver. We are also grateful to Dr. Mahesh Krishnamurthy and some others for their significant donations to us. Without all their financial support and help we would not have been able to attract the high level of scholarship that has contributed to this volume.

We are grateful to Bharatiya Vidya Bhavan (Bengaluru), IGNCA (Bengaluru), and The Mythic Society (Bengaluru), for helping us organizing the Swadeshi Indology Conference - 5. Bharatiya Vidya Bhavan allowed us to conduct the conference in its premises and ensured the highest level of professionalism in providing us the facilities. IGNCA, Bengaluru, took charge of the inauguration and the catering expenses for the event, and this was extremely timely and helpful for us to focus our attention to other things required for the conduct of this conference. In particular, we are thankful to Shri. H. N. Suresh of Bharatiya Vidya Bhavan who was available to us at any time both before and during the conference to ensure that nothing was amiss in the premises. His assurance and support were invaluable for us that helped us make a success of the event. Special thanks are due also to Dr. Deepthi Navaratna of IGNCA for the coordination she made possible with the IGNCA team to help us with the logistics and catering for the event. Sri V. Nagaraj of The Mythic Society has kindly agreed to take charge of the publication of the proceedings. The teams at these institutions put in enormous efforts to make the conference a success and we owe them a huge debt of gratitude for the same. 

