\thispagestyle{empty}

\begin{center}
\Huge\textbf{Acknowledgements}
\end{center}

\vskip 20pt

\begin{quote}
\textbf{\textit{“A hundred times I remind myself that my inner and outer life depends on the labour of other men, living and dead, and that I must exert myself in order to give in the same measure as I have received and am still receiving.”}}
\begin{flushright}
– Albert Einstein.
\end{flushright}
\end{quote}

This book is the handiwork of many. I am merely a tool who feels blessed to have been chosen by the Almighty for this purpose.

My son, Dr. Sooraj Tejaswi, is a gifted child for us. He has been\break an outstanding student throughout his career. Despite his busy\break career as the Director of Gastroenterology Fellowship Program, Asso\-ciate Professor of Gastroenterology, University of California, Davis,\break USA has played a key role in transforming the Kannada edition to its current English form. There could be no better honor for us as parents, than to have our own son help transform the Kannada edition, which is my brainchild.

World–renowned diabetologist, PadmaShri Dr. V. Mohan has crow\-ned our efforts by graciously penning the foreword for this book.

We are ever grateful to Justice Dr. M. Rama Jois, former Chief\break Justice of Punjab and Haryana High Court, and former Governor of Jharkhand and Bihar, who was magnanimous in reviewing this work and supporting the fight against diabetes.

This book is the result of a protracted \& vigorous collective effort.

“Behind every successful man there is a woman”, and that woman in my life is my wife, Prof. Lakshmidevi Narayan. She has supported me at every stage of my life, and has stood by me patiently as I burnt the midnight oil for this book.

My youngest daughter–in–law Mrs. Amrita Ranjan, wife of Dr.Sooraj Tejaswi was instrumental in designing the graphs for this book.

Dr. S. Venkat Rao(Late), nutrition scientist at the Central Food Technology Research Institute, Mysuru, worked with me for more than a decade on several scientific projects relating to diabetes, with which I gained experience and scientific background to venture into writing this book.

Members of the Ethics Committee of Shreehari Diabetes\break Foundation – Dr. B.K. Achar, Senior Physician and Chairman of the Committee, Mr.Rajashekara Koti(Late), distinguished journa\-list, socia\-list and winner of the highest award for journalism in Karnataka, the TSR award and Mr. B.K. Ramaiah, Senior Advocate, all played a vital role in the research projects designed and conducted at my center.

Dental problems are common in diabetics. My second son, Dr. Sunil Tejaswi, Associate Professor of Conservative Dentistry at J.S.S. Dental College, Mysore, has enriched the book (by offering his expertise on this topic), by authoring the chapter about dental problems in diabetes (Chapter \ref{chap18B}).

Mrs. Shantakumari. K, my office secretary patiently and expertly helped with computer–related work during the book–writing process. Mr.Gurunavalli helped with the illustrations in the book.

We are indebted to Shri B.V.Sridhar, Deputy Director of Prasaranga (Retired) and Editor of The Mysore University English–Kannada Dictionary and Editor of Oxford English–English Kannada Dictionary for his wholehearted involvement in proof reading this work.

We are grateful to Dr. George W. Meyer, former Governor of the American College of Physicians (ACP), Northern California Chapter, for proof reading this book and providing valuable suggestions in helping spread the message about diabetes.

My eldest son, Mr. Subhash Chandra, provided valuable inputs for the chapter on stress, yoga and meditation.

I am indebted to the efforts of many scientists and physicians who have contributed to the scientific literature on diabetes, without which, we could not have created this work.

Finally, over the past 50 years I have been blessed with the trust of my patients in my medical practice. Each of my patients opened new frontiers of scientific knowledge for me, and helped make me a better physician, and more importantly a better human being. Some of my dia\-betic patients gladly volunteered their experiences and photographs to embellish the book further. I am what I am because of my patients, and I am grateful to every single diabetic and their family members. Every single patient I have had the privilege to treat has been my guru, and enriched my knowledge, which has enabled me to write this book. I am extremely grateful to each one of them.

\begin{flushright}
\textbf{Dr. V. Lakshminarayan}
\end{flushright}

