
\chapter*{PUBLISHER'S NOTE}

\vskip .7cm

The Madhwa Samaj at Mambalam started over twenty-six years ago with the object of study, dissemination of knowledge of, propagation of Dwaita Siddhānta and tenets and precepts of Madhva cult among the general public and particularly among Mādhvas, has in its own humble way continued and continues to serve the ardent devotees of Mādhva faith. The Samaj has been conducting classes for study of works of Sri Madhva and his followers like Sri Jayathirtha, Sri Vādirāja, Sri Raghavendraswamy and others including Haridasas and has covered to some extent works on Prasthānatrayas, Harikathāmộtasāra etc. The Samaj also conducts Ārādhana Mahotsavas, celebrates Madhwa Navami, Dhātrihavana and such other religious functions. Inviting and paying respects to Peethadhipathis, Swamijis and invoking their blessings is another activity of the Samaj. Its main activity is conducting regular classes for the study of Madhwa Philosophy and it is a matter of some satisfaction that by the grace of Sri Hari and Sri Vāyu this has been going on uninterruptedly for the last twenty-six years i.e., since the inception of the Samaj. Special works like "Vishnurahasya', 'Sattattvaratnamāla' have also been covered in the studies.

Book publication on topics and works giving an elucidation of the deeper meanings therein, in simple Kannada, has also been one of the activities of the Samaj. Nārāyana Sabdārtha', 'Sriman Mahābhāratha Tātparya Nirņaya Bhāvasangraha', 'Sri Raghavendra Stotra' (Kannada meanings in Kannada script and Tamil script) Sri Rama Chāritra Manjari' are some of the publications by the Samaj.

The present book is titled 'Māghamāsa Māhātme' (in Kannada) by Sri A. R. Subbannachar-the patron of the Samaj and mainstay of all the activities of the Samaj (intellectual, spiritual, religious, study of dissemination of sāstric knowledge, publication etc.)

Māghamāsa Māhātme' is from Vāyu Purāņa-one of the Purāṇas elucidating Sāndilya Tattvas and Harimahātme and Hari Sarvottamatva. It is hoped that this publication will give the readers the knowledge of the greatness of Lord Vishnu and his Glory and Majesty and inculcate the spirit of deep devotion in Him.

This publication is rendered possible due to the munificent grant of financial assistance by 'Tirumala Tirupathi Devastanams' and we offer our deep-felt thanks to them.

Our thanks are due to Sri D. S. Krishnachar of Prabha Printing House, Bangalore-4, in getting this printed neatly unmindful of the difficulties.

\eject

Our Special thanks to Sri A. R. Subbannachar, the author of this work, for giving us the opportunity of publishing the work and helping us fulfil one of the objectives of our Samaj.

\begin{center}
\textbf{"Sri Krishnarpanamastu”}
\end{center}

\begin{flushright}
\textbf{R. Srinivasa Rao \textit{President, Madhva Samaj}}
\end{flushright}

\begin{flushleft}
13, Kuppaiah Street \\ West Mambalam \\ Madras-600033
\end{flushleft}

